---
title: "Analysis of U.S. Storm Event Data and the Impact on Public Health and the Economy"
author: "Jeffrey Hunter"
date: "4/24/2019"
output:
  html_document:
    df_print: paged
  pdf_document: default
  html_notebook: default
  pdf_notebook: default
---

## Table of Contents

[Introduction](#introduction)<br />
[Synonpsis](#synonpsis)<br />
[Set up environment](#set-up-environment)<br />
[Load Data](#load-data)<br />
[Data Processing](#data-processing)<br />
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Impact on Public Health](#impact-on-public-health)<br />
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Impact on Economy](#impact-on-economy)<br />
[Results](#results)<br />
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Event Types Most Harmful to Population Health](#event-types-most-harmful-to-population-health)<br />
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Event Types with Greatest Economic Consequences](#event-types-with-greatest-economic-consequences)<br />
[Conclusion](#conclusion)

## Introduction

*pending*

## Synonpsis

*pending*

## Set up environment

Disable warnings

```{r setup, include = TRUE}
knitr::opts_chunk$set(echo = FALSE)
```

Load libraries

```{r, echo = TRUE}
options(scipen = 1) # disable scientific notation for numbers
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
```

Session information

```{r, echo = TRUE}
sessionInfo()
```

## Load Data

Download data file from Internet (if it doesn't exit) and load the data.

```{r, echo = TRUE, cache = TRUE}
setwd("~jhunter/repos/coursera/reproducible-research-course-project-2")
stormDataFileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
stormDataFile <- "data/storm-data.csv.bz2"
if (!file.exists('data')) {
    dir.create('data')
}
if (!file.exists(stormDataFile)) {
    download.file(url = stormDataFileURL, destfile = stormDataFile)
}
# read.csv handles compressed files automatically
stormData <- read.csv(stormDataFile, sep = ",")
#activityData$date <- as.POSIXct(activityData$date, format = "%Y-%m-%d", tz = "")
```

Display dataset summary

```{r, echo = TRUE}
names(stormData)
```

```{r, echo = TRUE}
str(stormData)
```

```{r, echo = TRUE}
summary(stormData)
```

```{r, echo = TRUE}
head(stormData)
```

## Data Processing

### Impact on Public Health

*pending*

### Impact on Economy

*pending*

## Results

### Event Types Most Harmful to Population Health

*pending*

### Event Types with Greatest Economic Consequences

*pending*

## Conclusion

*pending*





