---
title: "Analysis of U.S. Storm Event Data and the Impact on Public Health and the Economy"
author: "Jeffrey Hunter"
date: "4/24/2019"
output:
  html_document:
    df_print: paged
  pdf_document: default
  html_notebook: default
  pdf_notebook: default
---

## Table of Contents

[Synonpsis](#synonpsis)<br />
[Set up environment](#set-up-environment)<br />
[Load Data](#load-data)<br />
[Data Processing](#data-processing)<br />
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Impact on Public Health](#impact-on-public-health)<br />
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Impact on Economy](#impact-on-economy)<br />
[Results](#results)<br />
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Event Types Most Harmful to Population Health](#event-types-most-harmful-to-population-health)<br />
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; [Event Types with Greatest Economic Consequences](#event-types-with-greatest-economic-consequences)<br />
[Conclusion](#conclusion)

## Synonpsis

This goal for this report is to analyze 
several weather events in the United States between the years 1950 and 2011



Storms and other severe weather events can cause both public health and economic
problems for communities and municipalities. Many severe events can result in
fatalities, injuries, and property damage, and preventing such outcomes to the
extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric
Administration's (NOAA) storm database. This database tracks characteristics of
major storms and weather events in the United States, including when and where
they occur, as well as estimates of any fatalities, injuries, and property
damage.


## Set up environment

Disable warnings

```{r setup, include = TRUE}
knitr::opts_chunk$set(echo = FALSE)
```

Load libraries

```{r, echo = TRUE}
options(scipen = 1) # disable scientific notation for numbers
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
```

Session information

```{r, echo = TRUE}
sessionInfo()
```

## Load Data

Download the compressed data file from the source URL (if not found locally) and
then load the compressed data file via `read.csv`. Prior to processing
the data, validate the downloaded data file and loaded dataset by checking the file 
size and dimensions respectively.

```{r, echo = TRUE, cache = TRUE}
setwd("~jhunter/repos/coursera/reproducible-research-course-project-2")
stormDataFileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
stormDataFile <- "data/storm-data.csv.bz2"
if (!file.exists('data')) {
    dir.create('data')
}
if (!file.exists(stormDataFile)) {
    download.file(url = stormDataFileURL, destfile = stormDataFile)
}
stormData <- read.csv(stormDataFile, sep = ",")
stopifnot(file.size(stormDataFile) == 49177144) 
stopifnot(dim(stormData) == c(902297,37))
```

Display dataset summary

```{r, echo = TRUE}
names(stormData)
```

```{r, echo = TRUE}
str(stormData)
```

```{r, echo = TRUE}
summary(stormData)
```

```{r, echo = TRUE}
head(stormData)
```

## Data Processing

### Impact on Public Health

*pending*

### Impact on Economy

*pending*

## Results

### Event Types Most Harmful to Population Health

*pending*

### Event Types with Greatest Economic Consequences

*pending*

## Conclusion

*pending*





